DEIXIS AND FRAMES OF REFERENCE IN DEDICATORY EPIGRAMS. The use of a database with an interdisciplinary approach

### 2015-12-18 source/42/Paper_licciardello.tex

\noindent In the Ancient Greek world it was customary to accompany a dedication to the gods with a – usually – short inscription.
This was normally chiselled on the dedicated object (or on its support) and it recorded the main information related to
the offering. The elements recurring in such inscriptions were the name of the dedicator, the verb of dedication and
the name of the god receiving the dedication.\footnote{ On archaic dedications, see \citep[ 1-14]{day_archaic_2010} for
an introduction to the genre.} The most common and widespread formula for dedications contained exactly these three
elements: \begin{otherlanguage*}{greek}ὁ δεῖνα ἀνέθηκε τῷ θεῷ\end{otherlanguage*}. A frequent variation was \begin{otherlanguage*}{greek}ὁ δεῖνά με ἀνέθηκε τῷ θεῷ\end{otherlanguage*}, \footnote{ The formula here presented (with its two variants) was \ identified by Maria Letizia Lazzarini, in
her study on the formulas of archaic dedications \citep[58-40]{lazzarini_formule_1976}. Her analysis is based on both verse and prose
inscriptions from the archaic age. However, such basic scheme, with these elements, was normal in epigrams as well and
it remained substantially unchanged in classical and post-classical time.} \ where the speaker is the object itself, as
is clear from the employment of the personal pronoun \begin{otherlanguage*}{greek}με\end{otherlanguage*}. It is interesting to note that since their first
examples the speaking object was a recurring feature in dedicatory epigrams and it is still frequent in the 4th century
BC.\footnote{ On the topos of the speaking object, see \citep{burzachechi_oggetti_1962}, \citep[36-52]{svenbro_phrasikleia:_1988}, \citep[16-27]{tueller_look_2008}, \citep[151-166]{furley_life_2010}, \citep[250-260]{wachter_origin_2010}, \citep[29-107]{christian_gebildete_2014}.} Clearly, the basic scheme here
presented could be varied by omitting some elements (as the name of the dedicator) or adding others (as the generic
name of the object, like the recurrent \begin{otherlanguage*}{greek}ἄγαλμα\end{otherlanguage*}).

A crucial moment in the evolution of the epigrammatic genre is the beginning of the Hellenistic era, when epigrams
began to be considered full literary creations and started to be circulated autonomously, no longer limited to one
inscription alone.\footnote{ This trend may be linked to the custom prevalent in the 5th and 4th century BC to copy
epigrams on monuments and then quote them in various texts. See \citep[47ff.]{gutzwiller_poetic_1998}} This development led to the
composition of literary epigrams not intended for inscription, but which in some way maintained the illusion of a
material inscription on a stone. The Hellenistic epigrammatists who worked with inscriptional type of epigrams (and
among these dedicatory epigrams) retained the structure, style and traditions of the epigraphic models, yet the
translation of these into a book context inevitably means that the communicative strategies employed until then had to
be reinvented. The primary reason for such changes is self-evident: the monument intended for inscription and the space
surrounding it do not exist anymore – they have to be imagined into reader's mind.\footnote{ This involvement of the
reader can be put in relation with what 
%\citep{Bing1995} 
defines Ergänzungsspiel, i.e. with the process of supplementation
of the text deliberately incited by Hellenistic epigrammatists.}
%reference to Bing 1995 not in biblio, asked to Licciardello

>>>>

\noindent In the Ancient Greek world it was customary to accompany a dedication to the gods with a – usually – short inscription.
This was normally chiselled on the dedicated object (or on its support) and it recorded the main information related to
the offering. The elements recurring in such inscriptions were the name of the dedicator, the verb of dedication and
the name of the god receiving the dedication.\footnote{ On archaic dedications, see \citep[spec. 1-14]{day_archaic_2010} for
an introduction to the genre.} The most common and widespread formula for dedications contained exactly these three
elements: \begin{otherlanguage*}{greek}ὁ δεῖνα ἀνέθηκε τῷ θεῷ\end{otherlanguage*}. A frequent variation was \begin{otherlanguage*}{greek}ὁ δεῖνά με ἀνέθηκε τῷ θεῷ\end{otherlanguage*}, \footnote{ The formula here presented (with its two variants) was \ identified by Maria Letizia Lazzarini, in
her study on the formulas of archaic dedications \citep[58-60]{lazzarini_formule_1976}. Her analysis is based on both verse and prose
inscriptions from the archaic age. However, such basic scheme, with these elements, was normal in epigrams as well and
it remained substantially unchanged in classical and post-classical time.} \ where the speaker is the object itself, as
is clear from the employment of the personal pronoun \begin{otherlanguage*}{greek}με\end{otherlanguage*}. It is interesting to note that since their first
examples the speaking object was a recurring feature in dedicatory epigrams and it is still frequent in the 4th century
BC.\footnote{ On the topos of the speaking object, see \citep{burzachechi_oggetti_1962}, \citep[36-52]{svenbro_phrasikleia:_1988}, \citep[16-27]{tueller_look_2008}, \citep[151-166]{furley_life_2010}, \citep[250-260]{wachter_origin_2010}, \citep[29-107]{christian_gebildete_2014}.} Clearly, the basic scheme here
presented could be varied by omitting some elements (as the name of the dedicator) or adding others (as the generic
name of the object, like the recurrent \begin{otherlanguage*}{greek}ἄγαλμα\end{otherlanguage*}).

A crucial moment in the evolution of the epigrammatic genre is the beginning of the Hellenistic era, when epigrams
began to be considered full literary creations and started to be circulated autonomously, no longer limited to one
inscription alone.\footnote{ This trend may be linked to the custom prevalent in the 5th and 4th century BC to copy
epigrams on monuments and then quote them in various texts. See \citep[47ff.]{gutzwiller_poetic_1998}} This development led to the
composition of literary epigrams not intended for inscription, but which in some way maintained the illusion of a
material inscription on a stone. The Hellenistic epigrammatists who worked with inscriptional type of epigrams (and
among these dedicatory epigrams) retained the structure, style and traditions of the epigraphic models, yet the
translation of these into a book context inevitably means that the communicative strategies employed until then had to
be reinvented. The primary reason for such changes is self-evident: the monument intended for inscription and the space
surrounding it do not exist anymore – they have to be imagined into reader's mind.\footnote{ This involvement of the
reader can be put in relation with what \citep{Bing1995} defines Ergänzungsspiel, i.e. with the process of supplementation
of the text deliberately incited by Hellenistic epigrammatists.}
